---
title: "Confidence Interval"
output: 
  prettydoc::html_pretty:
    theme: default
runtime: shiny
---

```{r, message=FALSE, include=FALSE, echo=FALSE}
library(tidyverse)

yield = read.csv("data/yield_df.csv")
```


Below is a plot of 20 random sample means and their confidence intervals, based on sets of 4 samples.  Confidence intervals that include the population mean are colored black; those that don't are colored red.  Press the "Click to Sample" several times and observe how many intervals include the population mean (the blue horizontal line).  Note: the plot may take a few seconds to render.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(shiny)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      actionButton("resample", "Click to Sample")
    ),
    mainPanel(
      plotOutput("plot")
    )
  )
)

server <- function(input, output, session) {
  observeEvent(input$resample, {
    output$plot = renderPlot({
      hist_list = list()
      for(i in c(1:20)){
        yield_sample = sample(yield$yield_bu, 4) %>%
            as.data.frame()
        plot_data = data.frame(sample_no = i)
        plot_data$no_of_samples = 4
        plot_data$sample_mean = mean(yield_sample$.)
        plot_data$std_dev = sd(yield_sample$.)
        plot_data$std_err = plot_data$std_dev/sqrt(4)
        plot_data$pop_mean = mean(yield$yield_bu)
        hist_list[[i]] = plot_data
        i = i+1
      }
      
      hist_df = do.call(rbind.data.frame, hist_list) %>%
        group_by(sample_no) %>%
        summarise(sample_mean = mean(sample_mean),
                  std_err = mean(std_err),
                  no_of_samples = mean(no_of_samples)) %>%
        ungroup() %>% 
        mutate(t_value = qt(0.975,3)) %>%
        mutate(upper_limit = sample_mean + t_value*std_err) %>%
        mutate(lower_limit = sample_mean - t_value*std_err) %>%
        mutate(error_bar_length = t_value*std_err) %>%
        mutate(pop_mean = mean(yield$yield_bu)) %>%
        mutate(color_name = if_else(upper_limit<pop_mean | lower_limit>pop_mean, "red", "black"))
      
      
      hist_df %>%
        ggplot(aes(x=sample_no, y=sample_mean, group=color_name)) +
        geom_point() +
        geom_errorbar(aes(x=sample_no, ymin=lower_limit, ymax=upper_limit, color=color_name)) +
        geom_hline(aes(yintercept=mean(yield$yield_bu)), color="blue") +
        # scale_color_manual(values = c("red", "black")) +
        scale_color_identity()
    })
  })
}

shinyApp(ui, server)


```

Again, both the standard error and the t-value we use for calculating the confidence interval decrease as the number of samples decrease, so the confidence interval itself will decrease as well.