---
title: "Sample Distribution"
output: 
  prettydoc::html_pretty:
    theme: default
runtime: shiny
---

```{r, message=FALSE, include=FALSE, echo=FALSE}
library(sf)
library(tidyverse)
library(rcompanion)
library(plotly)

yield = st_read("data/merriweather_yield_map/merriweather_yield_map.shp", quiet=TRUE)
```


The sample mean is an estimate of the true population mean.  The distribution around the sample mean describes not only the sample means, the range of possible values for the true mean.  I realize this is a fuzzy concept.  By studying the distribution of our sample values, we are able to describe the probability that the population mean is a given value.

To better understand this, lets look at an animation similar to the one above, except we have replaced the histogram with the sample distribution curve.  The animation shows how the distribution curve changes as the number of sampes (n in the equation above) increases.  The sample distribution curve is plotted in black.  The population mean is plotted as a vertical blue line.


```{r echo=FALSE}
library(shiny)

  hist_list = list()
  for(i in c(2:100)){
    set.seed(1776)
    yield_sample = sample(yield$yield_bu, i) %>%
        as.data.frame()
    hist_data = hist(yield_sample$., breaks = seq(from=55, to=105, by=5), plot = FALSE)
    plot_data = cbind(hist_data$mids, hist_data$density) %>%
      as.data.frame()
    names(plot_data) = c("midpoints", "density")
    plot_data$no_of_samples = i
    plot_data$sample_mean = mean(yield_sample$.)
    plot_data$std_dev = sd(yield_sample$.)
    plot_data$std_err = plot_data$std_dev/sqrt(i)
    plot_data$pop_mean = mean(yield$yield_bu)
    hist_list[[i]] = plot_data
    i = i+1
  }


  hist_df = do.call(rbind.data.frame, hist_list) %>%
  group_by(no_of_samples) %>%
  summarise(sample_mean = mean(sample_mean),
            std_err = mean(std_err))


x = seq(from=50, to=110, length=1000)

y_values = hist_df %>%
  group_by(no_of_samples) %>%
  do(y_val = dnorm(x, mean = .$sample_mean, sd = .$std_err)) %>%
  ungroup() %>%
  select(-no_of_samples) %>%
  unnest(cols=c(y_val))


x_values = hist_df %>%
  group_by(no_of_samples) %>%
  do(x_val = x) %>%
  unnest(cols=c(x_val))

plot_data = cbind(x_values, y_values) %>%
  as.data.frame()


ui = fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("no_samples", "Select number of samples", min = 2, max = 100, step=1, value = 2, 
                  animate = animationOptions(interval = 200))
    ),
    mainPanel(
      plotOutput("plot", height="300px")
    )
  )
)

server <- function(input, output, session) {
  output$plot = renderPlot({
    plot_data %>%
      filter(no_of_samples==input$no_samples) %>%
      ggplot(aes(x=x_val, y=y_val)) +
      geom_point(color="black") +
      geom_vline(aes(xintercept=mean(yield$yield_bu)), color = "blue") +
      lims(y=c(0,0.5))
  })

}

shinyApp(ui, server)
```



You should take away two important observations from this.  First, as the number of samples increases, the distribution curve narrows in width and increases in height.  Second, the sample distribtuion curve always includes the true population mean.  As a bonus, not that the curve in this animation is first centered about 15 bushels below the population mean and then approaches it as the sample number increases.    

It matters that the distribution curve narrows as the number of samples increases.  Why?  Because the narrower curve describes a narrower range of possible values for the population mean.  It increases our confidence that the population mean is close to the sample mean.

Also note as the number of samples increases from 2 to 10, the shape and location of the distribution curve changes more profoundly than it does as the number of samples increases, say, from 51 to 60.  If you take away nothing else from this lesson, understand whether you collect 2 or 3 samples has tremendous implications for your estimate of the population mean.  4 samples is much better than 3.  Do everything you can to fight for those first few samples. 