---
title: "Sampling from a Uniform Distribution"
output: 
  prettydoc::html_pretty:
    theme: default
runtime: shiny
---

```{r, message=FALSE, include=FALSE, echo=FALSE}
library(sf)
library(tidyverse)
library(rcompanion)
library(plotly)

yield = st_read("data/merriweather_yield_map/merriweather_yield_map.shp", quiet=TRUE)
```

What is even more powerful about the Central Limit Theorem is that our sample means are likely to be normally distributed, even if the population does not follow a perfect normal distribution.

Let's take this concept to the extreme.  Suppose we had a population where every value occurred with the same frequency.  This is known as a uniform distribution.  Then let's review this with the same app as above.



```{r, message=FALSE, echo=FALSE }
yield_uni = rep(seq(from=60, to=99.9, by=0.1), 15)
yield_uni_df = yield_uni %>%
  as.data.frame
names(yield_uni_df) = "yield_bu"

ui = fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("no_samples", "Select number of samples", min = 1, max = 100, step=1, value = 2)
    ),
    mainPanel(
      plotOutput("populationPlot", height="300px"),
      plotOutput("sampleDist", height="300px"),
      plotOutput("CLT", height="300px")
    )
  )
)

server = function(input, output){
  
  output$populationPlot = renderPlot({
    ggplot(yield_uni_df, aes(x=yield_bu)) +
      geom_histogram(fill="white", color="black", breaks=c(-Inf, seq(54.99,104.9, 2.5))) +
      lims(x=c(55, 105)) +
      geom_text(aes(x=100, y=300), label=paste0("SD = ", round(sd(yield_uni_df$yield_bu), 2)), size=8)
  })
  
  sampleData=reactive({
    set.seed(1995)
    sample_list = list()
    for(i in c(1:1000)){
      samples = mean(sample(yield$yield_bu, input$no_samples), na.rm = TRUE) %>%
        as.data.frame()
      sample_list[[i]] = samples
    }
    sample_list_df = do.call(rbind.data.frame, sample_list)
    names(sample_list_df) = "yield"
    sample_list_df
  })
  
  output$sampleDist = renderPlot({
    set.seed(1995)
    samples = sample(yield_uni_df$yield_bu, input$no_samples, replace=TRUE) %>%
        as.data.frame()   
    names(samples) = "yield"
    pop_mean = mean(yield_uni_df$yield_bu)
    sample_mean = mean(samples$yield)
    ggplot(samples, aes(x=yield)) +
      geom_histogram(fill="white", color="black") +
      geom_vline(xintercept = pop_mean, color = "red") +
      geom_vline(xintercept = sample_mean, color = "blue") +
      lims(x = c(55,105), y=c(0,14)) +
      geom_text(aes(x=100, y=7.5), label = paste0("\u03bc", " = ", round(pop_mean,1)), color="red", size=8) +
      geom_text(aes(x=100, y=6), label = paste0("\u03c3", " = ", round(sample_mean,1)), color="blue", size=8)
  })
  
  output$CLT = renderPlot({
    set.seed(2003)
    sample_list = list()
    for(i in c(1:1000)){
      samples = mean(sample(yield_uni_df$yield_bu, input$no_samples,replace=TRUE), na.rm = TRUE) %>%
        as.data.frame()
      sample_list[[i]] = samples
    }
    sample_list_df = do.call(rbind.data.frame, sample_list)
    names(sample_list_df) = "yield"
    pop_mean = mean(yield_uni_df$yield_bu)
    ggplot(sample_list_df, aes(x=yield)) +
      geom_histogram(fill="white", color="black") +
      geom_vline(xintercept = pop_mean, color = "red") +
      lims(x=c(55,105), y=c(0,600)) +
      geom_text(aes(x=100, y=300), label=paste0("SEM = ", round(sd(sample_list_df$yield), 2)), size=8)
      # lims(x=c(55, 105)) +
      # annotate(geom = "text", label = paste0("population mean = ", round(pop_mean,1)), x=Inf, y=Inf, vjust=0.5, color="red", size=8,  )
    
      # geom_text(aes(x=85, y=0.15), label = paste0("sample mean = ", round(sample_mean,1)), color="blue", size=8, hjust=0)
  })  
}

shinyApp(ui, server, options = list(height="1000px"))


```

We can see from the top plot that the distribution is uniform.  The distribution appears like a rectangle because all inividual values are occurring at the same frequency.

The middle plot again shows us the frequency of different sample values.  What do we notice as we increase the number of samples?  As it did for the normal distribution, the frequencies of sample values from the uniform distribution approximates that of the uniform population distribution from which they were drawn.  As the number of samples increases, their distribution becomes more rectangular.

But what happens in the bottom plot, which shows the distribution of samples?  As the number of samples per sample mean increases, the distribution of sample means narrows in width -- just the same as when the samples were pulled from the normally-distributed population.  When we take multiple samples, even from a uniform population, their mean will approach the population mean as the sample number increases.    